---
title: "Edmonton Topographic Map"
subtitle: "YEG-1 — Ring road extent + inner suburbs"
author: "Maps Project"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

Produce a layered static topographic map of Edmonton (ring road extent plus St. Albert, Sherwood Park, and Windermere) that combines:

-   **Terrain texture** — hillshade derived from a 19 m DEM
-   **Contour lines** — 10 m interval from the same DEM
-   **Physical features** — North Saskatchewan River, lakes, parks, green space
-   **Road network** — ring road, arterials (no residential clutter)

This serves as the foundation map for all subsequent `yeg-*` analyses.

# Data Genealogy

| Object        | Source                            | Notes                                           |
|------------------------|------------------------|------------------------|
| `dem`         | AWS Terrain Tiles via `{elevatr}` | z = 12, \~19 m/px                               |
| `hill`        | Derived from `dem`                | `terra::shade()`, azimuth 315°, angle 35°       |
| `contours`    | Derived from `dem`                | `terra::as.contour()`, 30 levels                |
| `rivers`      | OpenStreetMap via Geofabrik PBF   | `waterway` = river, stream, canal               |
| `water_polys` | OpenStreetMap via Geofabrik PBF   | `natural` = water                               |
| `parks`       | OpenStreetMap via Geofabrik PBF   | `leisure` = park, nature_reserve                |
| `green`       | OpenStreetMap via Geofabrik PBF   | `natural` = wood, scrub, grassland              |
| `roads`       | OpenStreetMap via Geofabrik PBF   | `highway` = motorway, trunk, primary, secondary |

CRS for all rendered layers: **EPSG:3400** (NAD83 / Alberta 10-TM Forest)

```{r}
#| label: reconcile-wd
#| echo: false
#| results: hide
#| message: false
#| warning: false
library(knitr)
opts_knit$set(root.dir = "../../")
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results = "show",
  message = FALSE,
  warning = FALSE,
  comment = NA,
  tidy    = FALSE,
  echo    = TRUE,
  fig.width  = 12,
  fig.height = 10,
  out.width  = "100%"
)
```

```{r}
#| label: load-packages
#| code-summary: "Load packages"
library(magrittr)
library(ggplot2)
library(dplyr)
library(sf)
library(terra)
library(elevatr)
library(osmextract)   # Geofabrik PBF -- no Overpass API rate limits
library(tidyterra)
library(ggspatial)

source("scripts/graphing/map-theme.R")
```

# Area of Interest

```{r}
#| label: define-bbox
#| code-summary: "Bounding box and CRS"
yeg_bbox <- c(xmin = -113.80, ymin = 53.38, xmax = -113.27, ymax = 53.72)
# elevatr requires an sf data frame (not bare sfc)
yeg_bbox_sf <- st_as_sf(st_as_sfc(st_bbox(yeg_bbox, crs = st_crs(4326))))
crs_ab <- 3400  # NAD83 / Alberta 10-TM Forest
```

Extent covers the Anthony Henday ring road with \~10 km margin to include St. Albert (NW), Sherwood Park (E), and Windermere (SW).

# Elevation Data

```{r}
#| label: download-dem
#| code-summary: "Download DEM and compute hillshade + contours"
#| cache: true
dem_raw <- elevatr::get_elev_raster(locations = yeg_bbox_sf, z = 12, src = "aws")
dem     <- terra::rast(dem_raw)
dem     <- terra::project(dem, paste0("EPSG:", crs_ab))

slope  <- terra::terrain(dem, "slope",  unit = "radians")
aspect <- terra::terrain(dem, "aspect", unit = "radians")
hill   <- terra::shade(slope, aspect, angle = 35, direction = 315)

contours <- terra::as.contour(dem, nlevels = 30) |>
  sf::st_as_sf() |>
  sf::st_transform(crs_ab)

# Map extent limits used in coord_sf to clip all layers to DEM bbox
map_ext <- terra::ext(dem)
xlims   <- c(map_ext$xmin, map_ext$xmax)
ylims   <- c(map_ext$ymin, map_ext$ymax)
```

# Physical Features (OSM)

```{r}
#| label: osm-queries
#| code-summary: "Read OSM layers from Geofabrik Alberta PBF (local, no API)"
#| cache: true
# PBF downloaded once to data-public/raw/osm/ via oe_get(); all reads are local.
osm_dir <- "data-public/raw/osm"
options(osmextract.download_directory = osm_dir)

oe_get("Alberta", provider = "geofabrik", layer = "lines",
       download_only = TRUE, quiet = TRUE, download_directory = osm_dir)
oe_get("Alberta", provider = "geofabrik", layer = "multipolygons",
       download_only = TRUE, quiet = TRUE, download_directory = osm_dir)

pbf_path <- normalizePath(
  list.files(osm_dir, pattern = "alberta.*\\.osm\\.pbf$", full.names = TRUE)[1]
)

clip_wgs <- yeg_bbox_sf
clip_ab  <- st_transform(clip_wgs, crs_ab)
safe_crop <- function(x, clip) {
  if (is.null(x) || nrow(x) == 0) return(x)
  st_crop(st_make_valid(x), clip)
}

read_lines <- function(query) {
  oe_read(pbf_path, layer = "lines", query = query,
          wkt_filter = st_as_text(st_geometry(clip_wgs)), quiet = TRUE) |>
    st_transform(crs_ab) |> safe_crop(clip_ab)
}
read_polys <- function(query) {
  oe_read(pbf_path, layer = "multipolygons", query = query,
          wkt_filter = st_as_text(st_geometry(clip_wgs)), quiet = TRUE) |>
    st_transform(crs_ab) |> safe_crop(clip_ab)
}

rivers      <- read_lines("SELECT * FROM lines WHERE waterway IN ('river','stream','canal')")
water_polys <- read_polys("SELECT * FROM multipolygons WHERE natural = 'water'")
parks       <- read_polys("SELECT * FROM multipolygons WHERE leisure IN ('park','nature_reserve')")
green       <- read_polys("SELECT * FROM multipolygons WHERE natural IN ('wood','scrub','grassland')")
roads       <- read_lines(
  "SELECT * FROM lines WHERE highway IN ('motorway','trunk','primary','secondary')"
) |> mutate(road_class = case_when(
    highway %in% c("motorway","trunk")    ~ "major",
    highway %in% c("primary","secondary") ~ "arterial",
    TRUE                                   ~ "other"
  ))
```

# Maps

## Terrain Layer (g1)

Hillshade and contour lines only — baseline terrain texture.

```{r}
#| label: g1-terrain
#| fig-cap: "Hillshade + 10 m contours derived from AWS Terrain Tiles z=12"
ggplot() +
  geom_spatraster(data = hill) +
  scale_fill_gradient(low = "grey15", high = "grey98",
                      guide = "none", na.value = "transparent") +
  geom_sf(data = contours, colour = "grey40", linewidth = 0.2, alpha = 0.6) +
  coord_sf(crs = crs_ab, xlim = xlims, ylim = ylims, expand = FALSE) +
  theme_map() +
  labs(
    title    = "Edmonton — Terrain",
    subtitle = "Hillshade + 10 m contours  |  DEM: AWS Terrain Tiles z=12",
    caption  = "CRS: NAD83 / Alberta 10-TM (EPSG:3400)"
  )
```

## Full Topographic Map (g2)

All layers combined.

```{r}
#| label: g2-full-map
#| fig-cap: "Layered topographic map of the Edmonton ring-road extent"
ggplot() +
  geom_spatraster(data = hill) +
  scale_fill_gradient(low = "grey15", high = "grey98",
                      guide = "none", na.value = "transparent") +
  geom_sf(data = green,       fill = "#c8dfc0", colour = NA, alpha = 0.6) +
  geom_sf(data = parks,       fill = "#b5d4a8", colour = NA, alpha = 0.7) +
  geom_sf(data = water_polys, fill = color_water, colour = NA) +
  geom_sf(data = rivers,      colour = color_water, linewidth = 0.5) +
  geom_sf(data = filter(roads, road_class == "major"),
          colour = "#d4945a", linewidth = 0.9) +
  geom_sf(data = filter(roads, road_class == "arterial"),
          colour = color_road, linewidth = 0.4) +
  geom_sf(data = contours, colour = "grey50", linewidth = 0.15, alpha = 0.4) +
  annotation_scale(location = "br", unit_category = "metric", text_cex = 0.7) +
  annotation_north_arrow(location = "tl",
                         style = north_arrow_minimal(text_size = 8)) +
  coord_sf(crs = crs_ab, xlim = xlims, ylim = ylims, expand = FALSE) +
  theme_map() +
  labs(
    title    = "Edmonton Metropolitan Area",
    subtitle = "Topographic reference map  |  Ring road extent + inner suburbs",
    caption  = "Sources: AWS Terrain Tiles, OpenStreetMap / Geofabrik  |  CRS: EPSG:3400"
  )
```

## Major Roads Only (g21)

Same as `g2` but arterial roads removed — cleaner at smaller print sizes.

```{r}
#| label: g21-major-roads
#| fig-cap: "Topographic map with major roads only (motorway + trunk)"
ggplot() +
  geom_spatraster(data = hill) +
  scale_fill_gradient(low = "grey15", high = "grey98",
                      guide = "none", na.value = "transparent") +
  geom_sf(data = green,       fill = "#c8dfc0", colour = NA, alpha = 0.6) +
  geom_sf(data = parks,       fill = "#b5d4a8", colour = NA, alpha = 0.7) +
  geom_sf(data = water_polys, fill = color_water, colour = NA) +
  geom_sf(data = rivers,      colour = color_water, linewidth = 0.5) +
  geom_sf(data = filter(roads, road_class == "major"),
          colour = "#d4945a", linewidth = 0.9) +
  geom_sf(data = contours, colour = "grey50", linewidth = 0.15, alpha = 0.4) +
  annotation_scale(location = "br", unit_category = "metric", text_cex = 0.7) +
  annotation_north_arrow(location = "tl",
                         style = north_arrow_minimal(text_size = 8)) +
  coord_sf(crs = crs_ab, xlim = xlims, ylim = ylims, expand = FALSE) +
  theme_map() +
  labs(
    title    = "Edmonton Metropolitan Area",
    subtitle = "Major roads only (motorway + trunk)  |  Ring road extent + inner suburbs",
    caption  = "Sources: AWS Terrain Tiles, OpenStreetMap / Geofabrik  |  CRS: EPSG:3400"
  )
```

## SW River Valley Zoom (g22)

Zoom onto the area around 5315 143 St NW — Terwillegar / Whitemud Creek confluence. Arterials restored at this scale for navigational context.

```{r}
#| label: g22-zoom-sw
#| fig-cap: "Zoom: SW Edmonton river valley around 5315 143 St NW"
zoom_bbox_wgs <- sf::st_bbox(
  c(xmin = -113.645, ymin = 53.440, xmax = -113.505, ymax = 53.545),
  crs = 4326
)
zoom_poly_ab <- sf::st_as_sfc(zoom_bbox_wgs) |> sf::st_transform(crs_ab)
zoom_ext     <- sf::st_bbox(zoom_poly_ab)
z_xlims      <- c(zoom_ext["xmin"], zoom_ext["xmax"])
z_ylims      <- c(zoom_ext["ymin"], zoom_ext["ymax"])

# 5315 143 St NW: ~53.490°N, 113.578°W
star_xy <- sf::st_sfc(sf::st_point(c(-113.578, 53.490)), crs = 4326) |>
  sf::st_transform(crs_ab) |>
  sf::st_coordinates()

ggplot() +
  geom_spatraster(data = hill) +
  scale_fill_gradient(low = "grey15", high = "grey98",
                      guide = "none", na.value = "transparent") +
  geom_sf(data = green,       fill = "#c8dfc0", colour = NA, alpha = 0.6) +
  geom_sf(data = parks,       fill = "#b5d4a8", colour = NA, alpha = 0.7) +
  geom_sf(data = water_polys, fill = color_water, colour = NA) +
  geom_sf(data = rivers,      colour = color_water, linewidth = 0.7) +
  geom_sf(data = filter(roads, road_class == "major"),
          colour = "#d4945a", linewidth = 1.2) +
  geom_sf(data = filter(roads, road_class == "arterial"),
          colour = color_road, linewidth = 0.5) +
  geom_sf(data = contours, colour = "grey50", linewidth = 0.2, alpha = 0.5) +
  annotate("text", x = star_xy[, "X"], y = star_xy[, "Y"],
           label = "\u2605", colour = "gold", size = 7) +
  annotation_scale(location = "br", unit_category = "metric", text_cex = 0.7) +
  annotation_north_arrow(location = "tl",
                         style = north_arrow_minimal(text_size = 8)) +
  coord_sf(crs = crs_ab, xlim = z_xlims, ylim = z_ylims, expand = FALSE) +
  theme_map() +
  labs(
    title    = "SW Edmonton \u2014 River Valley Zoom",
    subtitle = "\u2605 5315 143 St NW  |  Terwillegar / Whitemud area",
    caption  = "Sources: AWS Terrain Tiles, OpenStreetMap / Geofabrik  |  CRS: EPSG:3400"
  )
```

# Session Info

```{r}
#| label: session-info
#| code-summary: "Reproducibility snapshot"
cat("Report rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
sessionInfo()
```